---
title: "run_analysis"
author: "James Hamilton"
date: "March 5, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
```

# Data Wrangling exercise 3 - Human Activity recognition
Signals from the phone accelerometer were filtered by the experimenters and supplied in the X_test.txt and X_train.txt files while the activity performed is supplied in y_test.txt and y_train.txt. The subject identifier is in subject_test.txt and subject_train.txt.   

The following signals are provided:

Signal          | Signal            | Signal   
----------------|-------------------|-------------------
tBodyAcc-XYZ    | tGravityAcc-XYZ   | tBodyAccJerk-XYZ   
tBodyGyro-XYZ   | tBodyGyroJerk-XYZ | tBodyAccMag   
tGravityAccMag  | tBodyAccJerkMag   | tBodyGyroMag   
tBodyGyroJerkMag| fBodyAcc-XYZ      | fBodyAccJerk-XYZ   
fBodyGyro-XYZ   | fBodyAccMag       | fBodyAccJerkMag   
fBodyGyroMag    | fBodyGyroJerkMag   


'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

The set of variables that were estimated from these signals are:   

mean(): Mean value   
std(): Standard deviation   
mad(): Median absolute deviation   
max(): Largest value in array   
min(): Smallest value in array   
sma(): Signal magnitude area   
energy(): Energy measure. Sum of the squares divided by the number of values.   
iqr(): Interquartile range   
entropy(): Signal entropy   
arCoeff(): Autorregresion coefficients with Burg order equal to 4   
correlation(): correlation coefficient between two signals   
maxInds(): index of the frequency component with largest magnitude   
meanFreq(): Weighted average of the frequency components to obtain a mean frequency   
skewness(): skewness of the frequency domain signal   
kurtosis(): kurtosis of the frequency domain signal   
bandsEnergy(): Energy of a frequency interval within the 64 bins of the FFT of each    window.
angle(): Angle between to vectors.   

Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:  

gravityMean, tBodyAccMean, tBodyAccJerkMean, tBodyGyroMean, tBodyGyroJerkMean

## 1 - Merge the training and the test sets to create one data set.

First merge the measurement data sets (test and train). I pull in the features first b/c it's easy with readr package to assign the column names while loading the data.

**Note: I had to pre-clean the X_test.txt and X_train.txt files in the shell because the "readr" package did not ignore the leading spaces in the line and treated multiple spaces as empty fields. The data set also had an extraneous \r at the end of the lines.**

I also assigned column names for the "X" dataset here using the feature names pulled from "features.txt".
```{r}
activities <- read_delim("UCI HAR Dataset/activity_labels.txt", " ",
              col_names = c("ActivityLabel", "ActivityName"))
features <- read_delim("UCI HAR Dataset/features.txt",
                       " ", col_names = FALSE)
cnames <- make.names(features$X2, unique = TRUE)
x_tbl <- read_delim("UCI HAR Dataset/train/X_train.txt", 
                   " ", col_names = cnames)
x_tbl <- bind_rows(
  x_tbl, read_delim("UCI HAR Dataset/test/X_test.txt",
                  " ", col_names = cnames))
```

Merge the activity labels from the test & train data sets
```{r}
y_tbl <- read_delim("UCI HAR Dataset/train/y_train.txt",
                   " ", col_names = c("ActivityLabel"))
y_tbl <- bind_rows(
  y_tbl, read_delim("UCI HAR Dataset/test/y_test.txt",
          " ", col_names = c("ActivityLabel")))
```

Now merge the subjects into one tbl
```{r}
subject_tbl <- read_csv("UCI HAR Dataset/train/subject_train.txt",
                        col_names = c("subjectid"))
subject_tbl <- bind_rows(
  subject_tbl, read_csv("UCI HAR Dataset/test/subject_test.txt",
                        col_names = c("subjectid")))
```

## 2 - Extracts columns containing mean and standard deviation for each measurement (Hint: Since some feature/column names are repeated, you may need to use the make.names() function in R)

Use select to pull all required columns - column names were assigned when the data was loaded.
```{r}
mean_n_std <- x_tbl %>% select(contains("mean"), contains("std"))
```

## 3 - Creates variables called ActivityLabel and ActivityName that label all observations with the corresponding activity labels and names respectively.
Joined y_tbl with activities and then mash subject_tbl, y_tbl, and x_tbl together with bind_cols.   
Finally used gather achieve tidiness by moving the features to rows. "FeatureName" is the name of the measurement from the features file and value is the measurement listed for that column in the x_tbl.
```{r}
y_tbl <- left_join(y_tbl, activities, by = "ActivityLabel")
tidydata <- bind_cols(subject_tbl, y_tbl, x_tbl)
tidydata <- gather(tidydata, "FeatureName", "Value", 4:564)
rm(x_tbl, y_tbl, subject_tbl)
```

## 4 - From the data set in step 3, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

Using the pipe operatior I just grouped by subject, activity and feature and then summarised the feature measurement using "mean".
```{r}
tidysummary <- tidydata %>%
  group_by(., subjectid, ActivityName, FeatureName) %>%
  summarise(., AverageValue = mean(Value))
```

## 5 - Output the summary table
```{r}
write.csv(tidysummary,
          file = "human_activity_summary.csv",
          row.names = FALSE)
```

